Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-25 overleaf/introduction.tex

MPC theory dates back to the early 1980-ies~\cite{FOCS:Yao82b,STOC:GolMicWig87,STOC:BenGolWig88,STOC:ChaCreDam88}.
Long the realm of theoretical cryptography, MPC has seen significant advances in programming technology in recent years.
These advances bring MPC closer to practice and wider applicability ---
MPC technology has been employed in real-world scenarios such as auctions~\cite{FC:BCDGJK09}, biometric identification~\cite{ESORICS:BlaGas11},
and privacy-preserving machine learning~\cite{SP:MohZha17,CCS:MohRin18}.
The goal is to improve technology so that programmers can write \emph{secure} and \emph{efficient} programs without
commanding extensive knowledge of cryptographic primitives.

The problem, therefore, is to build a high-level programming language and a compiler, and there has been significant advance in
this space, e.g., ~\cite{CCS:BenNisPin08,ESORICS:BogLauWil08,CCS:ZhaSteBla13,Hastings:2019,Keller:2020,Acay:2021,Braun:2022} among other work.
Current research largely falls at the two ends
of the classical compiler: (1) work on \emph{front-end} language design and (2) work on \emph{back-end} protocol implementation. Work on language design
focuses on high-level constructs necessary to express multiple parties, computation by different parties, and information flow from one party
to another~\cite{SP:RasHamHic14, Acay:2021}. On the other end, work on protocol implementation focuses on cryptographic foundations and their efficient circuit-level
implementation~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}, e.g., implementation of operations (e.g., MUL, ADD) using different sharing protocols
(Boolean or Arithmetic GMW~\cite{STOC:GolMicWig87} or Yao's garbled circuits~\cite{FOCS:Yao82b}), as well as efficient share conversion from one representation
to another.

@@

We formalize the MPC Source~\cite{Ishaq:2019} intermediate representation and emphasize optimization over MPC Source. As in classical compilers, we envision different front ends (e.g., our front end IMP Source) compiling into MPC Source. MPC Source is particularly suitable for optimizations such as protocol mixing~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022} and SIMD-vectorization,
which takes advantage of amortization at the circuit level. The MPC Source IR exposes the \emph{linear structure} of MPC programs, which simplifies program analysis; this is
in contrast to source, which has if-then-else constructs. In the same time, MPC Source is sufficiently ``high-level''  to support analysis and optimizations that take into account
control and data flow in a specific program. %MPC Source facilitates cost modeling. 
%MPC Source is small in size and analysis is tractable, as opposed to analysis over an unrolled circuit~\cite{Ishaq:2019}. 
Again as in classical compilers, we envision translation of MPC Source (optimized or unoptimized) into MOTION, SPDZ, or other back-end code.

@@

We focus on the MOTION framework as our back-end for several reasons. First, it demonstrates high performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations (e.g., MUL, CMP, etc.) in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations (e.g., MUL\_SIMD), which amortize cost and lead to significant improvement in memory footprint and throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It enables MPC Source-level vectorization, a key focus of this paper.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model, which essentially minimizes the length of the schedule) was tractable, as the problem appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. Cost modeling is important as it drives not only vectorization but optimizations such as protocol mixing and scheduling as well~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}.

@@

Our results emphasize the importance of backend-independent optimizations --- vectorization (described in this work) and protocol mixing (tackled in previous works~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}) are two optimizations readily available at the level of MPC Source. We believe that our work can lead to future work on backend-independent compilation and optimization, ushering new MPC optimizations and combinations of optimizations in the vein of standard compilers, and thus bringing MPC programming technology closer to practice and wider applicability.

>>>>

MPC theory dates back to the early 1980-ies~\cite{Yao:1982,Goldreich:1987,Ben-Or:1988,Chaum:1988}.
Long the realm of theoretical cryptography, MPC has seen significant advances in programming technology in recent years.
These advances bring MPC closer to practice and wider applicability ---
MPC technology has been employed in real-world scenarios such as auctions~\cite{Bogetoft:2009}, biometric identification~\cite{Blanton:2011},
and privacy-preserving machine learning~\cite{Mohassel:2017,Mohassel:2018}.
The goal is to improve technology so that programmers can write \emph{secure} and \emph{efficient} programs without
commanding extensive knowledge of cryptographic primitives.

The problem, therefore, is to build a high-level programming language and a compiler, and there has been significant advance in
this space, e.g., ~\cite{Ben-David:2008,Bogdanov:2008,Zhang:2013,Hastings:2019,Keller:2020,Acay:2021,Braun:2022} among other work.
Current research largely falls at the two ends
of the classical compiler: (1) work on \emph{front-end} language design and (2) work on \emph{back-end} protocol implementation. Work on language design
focuses on high-level constructs necessary to express multiple parties, computation by different parties, and information flow from one party
to another~\cite{Rastogi:2014, Acay:2021}. On the other end, work on protocol implementation focuses on cryptographic foundations and their efficient circuit-level
implementation~\cite{Demmler:2015, Araki:2018, Braun:2022}, e.g., implementation of operations (e.g., MUL, ADD) using different sharing protocols
(Boolean or Arithmetic GMW~\cite{Goldreich:1987} or Yao's garbled circuits~\cite{Yao:1982}), as well as efficient share conversion from one representation
to another.

@@

We formalize the MPC Source~\cite{Ishaq:2019} intermediate representation and emphasize optimization over MPC Source. As in classical compilers, we envision different front ends (e.g., our front end IMP Source) compiling into MPC Source. MPC Source is particularly suitable for optimizations such as protocol mixing~\cite{Buscher:2018b,Ishaq:2019, Fang:2022} and SIMD-vectorization,
which takes advantage of amortization at the circuit level. The MPC Source IR exposes the \emph{linear structure} of MPC programs, which simplifies program analysis; this is
in contrast to source, which has if-then-else constructs. In the same time, MPC Source is sufficiently ``high-level''  to support analysis and optimizations that take into account
control and data flow in a specific program. %MPC Source facilitates cost modeling. 
%MPC Source is small in size and analysis is tractable, as opposed to analysis over an unrolled circuit~\cite{Ishaq:2019}. 
Again as in classical compilers, we envision translation of MPC Source (optimized or unoptimized) into MOTION, SPDZ, or other back-end code.

@@

We focus on the MOTION framework as our back-end for several reasons. First, it demonstrates high performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations (e.g., MUL, CMP, etc.) in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Buscher:2018b,Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations (e.g., MUL\_SIMD), which amortize cost and lead to significant improvement in memory footprint and throughput~\cite{Demmler:2015, Araki:2018, Braun:2022}. It enables MPC Source-level vectorization, a key focus of this paper.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model, which essentially minimizes the length of the schedule) was tractable, as the problem appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. Cost modeling is important as it drives not only vectorization but optimizations such as protocol mixing and scheduling as well~\cite{Buscher:2018b,Ishaq:2019, Fang:2022}.

@@

Our results emphasize the importance of backend-independent optimizations --- vectorization (described in this work) and protocol mixing (tackled in previous works~\cite{Buscher:2018b,Ishaq:2019, Fang:2022}) are two optimizations readily available at the level of MPC Source. We believe that our work can lead to future work on backend-independent compilation and optimization, ushering new MPC optimizations and combinations of optimizations in the vein of standard compilers, and thus bringing MPC programming technology closer to practice and wider applicability.

### 2022-05-25 overleaf/overview.tex

(a) IMP Source & (b) MPC Source & (c) Optimized MPC Source
\end{tabular}
\caption{Biometric Matching: ==== From (a) IMP Source to (b) MPC Source: 
First, MPC Source is an SSA form.
Second, it is linear. The conditional in lines 13-14 in IMP Source turns into the linear code in lines 12-16 in MPC Source.
The test turns into the CMP operation {\sf t = CMP(sum!3,min\_sum!2)}, followed by the
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of {\sf min\_sum}: if {\sf t} is true, then {\sf min\_sum} gets the value of the second multiplexer
argument,  {\sf min\_sum!3}, otherwise it takes the value of the third argument, {\sf min\_sum!2}.
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into
MUX operations, while the $\phi$-nodes at for-loops turn into \emph{pseudo} PHI nodes with a straightforward semantics.
==== From (b) MPC Source to (c) Optimized MPC Source: 
The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)).
Notation {\sf p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension.
The loop in lines 12-16 in (c) illustrates; here {\sf p[I,j]} refers to the $j$-th column in {\sf p}.
Unfortunately, CMP and MUX remain sequential.
% \ana{Add explanation to caption, this is hard to see.}
}
\label{tab:source_and_MPC_source_and_optimized_MPC_source}
\end{table*}

@@

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication (lines 9-10 in (c)), the second term corresponds to the for-loop on $j$ (lines 12-16) and the third
one corresponds to the remaining for-loops on $i$ (lines 19-25). 
Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe that (1) $\alpha_\mathit{var} \approx 0$ and (2) there is orders of magnitude improvement in running time and memory. 
E.g., we see about 12x improvement in online time in GMW for $N=128$. Additionally, the non-vectorized version runs out of memory for $N=256$, 
while the vectorized one runs with the standard maximal input size $N=4,096$.
%\ana{Edit numbers with final experiments.}

>>>>

(a) IMP Source & (b) MPC Source & (c) Optimized MPC Source
\end{tabular}
\caption{Biometric Matching: ==== From (a) IMP Source to (b) MPC Source: 
First, MPC Source is an SSA form.
Second, it is linear. The conditional in lines 13-15 in IMP Source turns into the linear code in lines 12-16 in MPC Source.
The test turns into the CMP operation {\sf t = CMP(sum!3,min\_sum!2)}, followed by the
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of {\sf min\_sum}: if {\sf t} is true, then {\sf min\_sum} gets the value of the second multiplexer
argument,  {\sf min\_sum!3}, otherwise it takes the value of the third argument, {\sf min\_sum!2}.
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into
MUX operations, while the $\phi$-nodes at for-loops turn into \emph{pseudo} PHI nodes with a straightforward semantics.
==== From (b) MPC Source to (c) Optimized MPC Source: 
The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)).
Notation {\sf p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension.
The loop in lines 12-16 in (c) illustrates; here {\sf p[I,j]} refers to the $j$-th column in {\sf p}.
Unfortunately, CMP and MUX remain sequential.
% \ana{Add explanation to caption, this is hard to see.}
}
\label{tab:source_and_MPC_source_and_optimized_MPC_source}
\end{table*}

@@

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication (lines 9-10 in (c)), the second term corresponds to the for-loop on $j$ (lines 12-16) and the third
one corresponds to the remaining for-loops on $i$ (lines 19-25). 
Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe that (1) $\alpha_\mathit{var} \approx 0$ and (2) there is orders of magnitude improvement in running time and memory. 
E.g., we see about 12x improvement in online time in GMW for $N=128$. Additionally, the non-vectorized version runs out of memory for $N=256$, 
while the vectorized one runs with the standard maximal input size $N=4,096$ (and perhaps larger $N$).
%\ana{Edit numbers with final experiments.}

### 2022-05-25 overleaf/prob_stmt_np_hardness.tex

Our goal is to construct a parallel schedule $P$ that reduces the program cost (when compared to cost of $S$), possibly an optimal schedule. 
Originally we hoped that the problem is simper and computation of the optimal schedule is tractable. Unfortunately, the optimal schedule turns out to be NP-hard via a reduction 
to the Shortest Common Supersequence problem.

@@

%The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem~\cite{Vazirani:2010}. 
The shortest common supersequence problem~\cite{Vazirani:2010} is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. %For our problem $n$ is the maximum length of a sequence and $k$ is the number of instructions of sequences. 
We can see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

>>>>

Our goal is to construct a parallel schedule $P$ that reduces the program cost (when compared to cost of $S$), possibly an optimal schedule. 
Originally we hoped that the problem is simper and computation of the optimal schedule was tractable. Unfortunately, the optimal schedule turns out to be NP-hard via a reduction 
to the Shortest Common Supersequence problem.

@@

%The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem~\cite{Vazirani:2010}. 
The shortest common supersequence problem~\cite{Vazirani:2010} is as follows: {\it given two or more sequences find the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. %For our problem $n$ is the maximum length of a sequence and $k$ is the number of instructions of sequences. 
We can see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

### 2022-05-25 overleaf/tech_report/compiler.tex

\figref{fig:compiler_framework} presents an overview of our compiler.
%We start the section with a brief description of the top-level syntax and semantic restrictions in~\secref{sec:syntax}.
%We proceed to describe the front end of the compiler:~\secref{sec:imp_to_ssa}
%describes translation from IMP Source to SSA, and~\secref{sec:ssa_to_MPC_Source}
%describes translation form SSA into MPC Source.
This section outlines the front end, \secref{sec:vectorization} describes analysis on MPC Source and backend-independent vectorization 
and~\secref{sec:backend} outlines the back end. 
%We describe analysis on MPC Source and backend-independent vectorization in~\secref{sec:vectorization}
%and translation into MOTION code in~\secref{sec:backend}. 
Details on all parts appear in~\cite{Anon_TR}.
%\ana{Need to state this more precisely.}

@@

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that computation is bounded~\cite{Demmler:2015,Buscher:2018b}.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[$i,j,k$] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[$i,j+2$] = ...} is not allowed. (This is again a restriction we imposed for convenience in our current implementation; we plan to extend the compiler with arbitrary write access.)
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}

\begin{comment}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{comment}

\end{enumerate}

\begin{comment}
The arguments of MUX are restricted to base types. 
(Again, this is just a restriction of our current implementation that will be lifted.)
This causes an inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

%\end{comment}

We note that while the programmer writes
annotations at the level of IMP Source (as in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a)), the
annotations propagate through the transformations; annotations are inferred and checked with a taint analysis at the 
level of MPC Source. The programmer annotates only inputs.
\end{comment}

@@

Our compiler translates from Source to SSA as sketched below. %Full details are in the extended version~\cite{Anon_TR}.

%\begin{enumerate}
%  \item
        \squeeze
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \squeeze
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        %This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \squeeze
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}
        Convert the restricted-syntax AST to a three-address control-flow graph (CFG). The step processes for-loops, if-statements and assignments as restricted by the syntax.
\begin{comment}
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
\end{comment}        
  %\item
        \squeeze
        \squeeze
        \paragraph{SSA conversion:}
        Convert 3-address CFG to SSA with Cytron's algorithm~\cite{Cytron:1991}.
%\end{enumerate}

@@

Once the compiler converts the code to SSA,
it transforms $\phi$-nodes that correspond to if-statements into MUX nodes.
From the 3-address CFG conversion step,
$\phi$-nodes corresponding to if-statements will be in a basic block
with the ``merge condition'' property.
E.g., if {\sf X!3 = $\phi$(X!1,X!2)} is in a block with merge condition {\sf C},
the compiler transforms it into {\sf X!3 = MUX(C, X!1, X!2)}.
Next, the compiler runs the dead code elimination algorithm from~\cite{Cytron:1991}. %Cytron's SSA paper.

Next, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
%\begin{comment}
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
(The Python AST naturally gives rise to a translation where each conditional has exactly two targets,
and each ``merge condition'' block has exactly two incoming edges, a TRUE and a FALSE edge.
Thus, each $\phi$-node has exactly two multiplexer arguments, which dovetails into MUX.
This is in contrast with Cytron's algorithm which operates at the level of the CFG and allows for
$\phi$-nodes with multiple arguments.)
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends to the main MPC source.
%\end{comment}

>>>>

\figref{fig:compiler_framework} presents an overview of our compiler.
We start the section with a brief description of the top-level syntax and semantic restrictions in~\secref{sec:syntax}.
We proceed to describe the front end of the compiler:~\secref{sec:imp_to_ssa}
describes translation from IMP Source to SSA, and~\secref{sec:ssa_to_MPC_Source}
describes translation form SSA into MPC Source.

We describe analysis on MPC Source and backend-independent vectorization in~\secref{sec:vectorization}
and translation into MOTION code in~\secref{sec:backend}.
%\ana{Need to state this more precisely.}

@@

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[$i,j,k$] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[$i,j+2$] = ...} is not allowed. (This is again a restriction we imposed for convenience in our current implementation; we plan to extend the compiler with arbitrary write access.)
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{enumerate}
The arguments of MUX are restricted to base types. 
(Again, this is just a restriction of our current implementation that will be lifted.)
This causes an inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes
annotations at the level of IMP Source (as in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a)), the
annotations propagate through the transformations; annotations are inferred and checked with a taint analysis at the 
level of MPC Source. Tthe programmer annotates only inputs.

@@

Our compiler translates from Source to SSA as follows:

%\begin{enumerate}
%  \item
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

        Convert the restricted-syntax AST to a three-address control-flow graph.
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
  %\item
        \paragraph{SSA conversion:}
        Convert the 3-address CFG to SSA with Cytron's algorithm.
%\end{enumerate}

@@

Once the compiler converts the code to SSA,
it transforms $\phi$-nodes that correspond to if-statements into MUX nodes.
From the 3-address CFG conversion step,
$\phi$-nodes corresponding to if-statements will be in a basic block
with the merge condition property.
For example, if {\sf X!3 = $\phi$(X!1,X!2)} is in a block with merge condition {\sf C},
the compiler transforms it into {\sf X!3 = MUX(C, X!1, X!2)}.
Next, the compiler runs the dead code elimination algorithm from Cytron's SSA paper.

Next, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
(The Python AST naturally gives rise to a translation where each conditional has exactly two targets,
and each ``merge condition'' block has exactly two incoming edges, a TRUE and a FALSE edge.
Thus, each $\phi$-node has exactly two multiplexer arguments, which dovetails into MUX.
This is in contrast with Cytron's algorithm which operates at the level of the CFG and allows for
$\phi$-nodes with multiple arguments.)
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends these to the main MPC source.
